Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Overall proofreading and editing of conclusions

  \begin{abstract}
  In this paper we analyze the changes in accessibility to food banks
  and related services before and during the COVID-19 pandemic in the
  City of Hamilton, Ontario. Food banks and services are the last line
  of support for households facing food insecurity; as such, they have
  become essential and their relevance cannot be ignored in the midst of
  the economic upheaval caused by the pandemic. Our analysis is based on
  the application of balanced floating catchment areas, and concentrates
  on households with lower incomes (\$\textless CAD40,000, approximately
  the Low Income Cutoff Value for a city of Hamilton's size).
  Accessibility was low to begin with in suburban and exurban parts of
  the city. We find that about 14\% of locations originally available in
  Hamilton closed during the pandemic; the impact of those closures on
  the level of service and accessibility was disproportionate, with
  system-wide losses exceeding 40\%. Those losses were geographically
  and demographically uneven. While every part of the city faced a
  reduction in accessibility, inner suburbs fared worse in terms of loss
  of accessibility. As well, children (age \(\le 18\)) appear to have
  been impacted the most.
  \end{abstract}

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and physical and mental health in children and adults (Elgar et al.,
2021; Jones, 2017; Ramsey et al., 2011; Seligman et al., 2010; Stuff et
al., 2004). Over at least the past four decades food banks and related
services have become an essential line of defense against food
insecurity in Canadian communities (Black and Seto, 2020; Holmes et al.,
2018; Riches, 2002; Tarasuk et al., 2020). In this respect, Canada is
not unlike numerous other wealthy countries where a systematic
dismantling of the welfare state took place in the intervening period
(Tarasuk et al., 2014).

The emergence of COVID-19, the worst public health crisis since the 1918
flu pandemic, has revealed important social and economic fault lines,
and pre-existing patterns of inequality appear to have been exacerbated.
Along several other dimensions of stress (e.g., accessibility to health
care facilities, Pereira et al., 2021a), this seems to be the case for
food insecurity as well (Laborde et al., 2020). In the US, for example,
it has been estimated that there was an increase of more than 30\% in
household food insecurity, and more than one third of households were
discovered to be newly food insecure - meaning they did not experience
food insecurity before the pandemic (Niles et al., 2020). In Canada, Men
and Tarasuk (2021) report that about 25\% of individuals who experienced
job insecurity (a relatively common occurrence during the pandemic),
also experienced food insecurity. Similarly, according to Statistics
Canada (2020a), in the early stages of the pandemic almost 15\% of
individuals reported living in a household that faced food insecurity;
the risk of food insecurity was substantially higher for households with
children. The difference between households with and without children
was significant, and 11.7\% of households with children indicated that
``food didn't last and {[}there was{]} no money to get more'' sometimes
or often, compared to 7.3\% of households without children; likewise,
13\% of households with children indicated that they ``{[}c{]}ouldn't
afford balanced meals'' sometimes or often, compared to 8.8\% of
households without children.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While food banks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As recently as 2019, the Hamilton
Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that food banks in Hamilton, Ontario, recorded the highest number
of visitors in the past 29 years; the number of children visiting food
banks (minors up to 18 years old) was 9,125 in March 2019, up from 8,278
the year before, a rate of increase greater than population growth.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary, which suggests that many
households may again turn to food bank services to meet their needs.

@@

\textbf{Need to provide some context here: the geography which makes
transportation challenging, i.e., due to the escarpment; the main parts
of the city: city core, inner suburbs, industrial north, suburban,
exurban; and the industrial background of the city maybe, with poverty
rates compared to other Canadian cities?}

@@

The locations of food banks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. Table 1 defines each
service type and the number of locations pre- and during the COVID-19
pandemic. While some food bank services have a specific a target
population, such as prioritizing families with young children aged
between 0 and 3 or accepting only those providing proof of low-income
status through housing and utility statements, all the food bank
services indicated below are designed to accommodate those in need of
food at zero to low cost. With our focus on food banks and food services
that offer free or low-cost meals at particular locations, we first
removed services such as Meals on Wheels and other food access services
such as food box, community kitchens, student nutrition programs, and
shopping and transportation. In addition, two free meal services held on
different days at the same location were collapsed into a single service
point for the accessibility analysis. Additional details on the
operations of individual facilities is not publicly available, and with
the changes in operations it proved unfeasible to collect it. For this
reason, the analysis to follow is of accessibility to the location of
food banks and services, but not to specific services (e.g., breakfasts
vs.~food boxes).

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: while increases in
accessibility that result from decreases in the travel time window might
seem counter-intuitive, this is a result of lower \emph{congestion},
since fewer households are serviced and thus competition for the same
resources is more limited. System-wide accessibility declines with
higher values of \(\delta\): as more households are serviced, congestion
grows and the level of service declines, although this happens at a
declining rate. We are not aware of any research that explains how long
people are expected to travel for food banks, but we note that in
developing countries, access to drinking water is defined as a source
that takes less than 30 minutes to reach (round trip, see UNICEF-WHO,
2019). There is no reason why people in affluent countries should be
expected to travel more for a basic need such as food. Accordingly, we
adopt a 15-minute threshold for the analysis (representing a one-way
trip). This threshold is also approximately where the rate of change in
accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially
during the pandemic, with accessibility dropping on average by almost
41\%, but with large variations: some zones experienced changes in
accessibility of only about 10\%, whereas the most affected zone saw a
loss of accessibility of almost 71\%. Figure
\ref{fig:accessibility-changes-with-local-i} shows the changes in
accessibility. Every zone is worse off after the closure of facilities
during the pandemic, but some parts of the city seem to have been
particularly affected. To better highlight these changes, we used a
local indicator of spatial autocorrelation (Anselin, 1995) to explore
the pattern of change in accessibility. Twenty-four TAZ are flagged as
having significant large losses of accessibility (at \(p<=0.10\),
without correcting for multiple comparisons). Those zones are
highlighted in the figure, where it can be seen that they form more or
less compact neighborhoods. Remarkably, the largest (and significant)
drops in accessibility are not downtown, but located in two cases in the
industrial north of the city, in one case in an inner suburb above the
escarpment, and lastly in a more suburban/exurban region in the
south-west.

@@

Words go here.

We use data from the census to model low-income households as a proxy
for households that are possibly experiencing food insecurity. However,
the degree to which our low-income cut-off of CAD\$40,000 reflects food
insecurity in the different zones of the study area is not known.

While the travel survey allows us to model multi-modal accessibilities
that align with travel behaviour observed in the travel survey and
capture differences in accessibility by age categories, the use of
travel survey data for modelling food insecurity also has its
limitations. Research into the population weighting methods used in the
TTS note that the survey may under-count the lowest- and highest-income
households in the survey study region, although the magnitude of this
under-counting is unknown as approximately 20\% of respondents to the
2016 survey did not report their income (Rose, 2018). In that regard,
the modal splits of low-income households observed through the TTS data
may not accurately reflect the travel behaviours of food insecure
populations.

In the absence of information on how food insecure households travel to
food banks and related services, we adopted a normative approach that
examines accessibility to food banks reachable within 15 minutes based
on standards for accessible drinking water\ldots{}

@@

@@

\leavevmode\hypertarget{ref-delamater2013spatial}{}%
Delamater, P.L., 2013. Spatial accessibility in suboptimally configured
health care systems: A modified two-step floating catchment area
(M2SFCA) metric. Health \& Place 24, 30--43.
doi:\href{https://doi.org/10.1016/j.healthplace.2013.07.012}{10.1016/j.healthplace.2013.07.012}

@@

@@

@@

@@

>>>>

  \begin{abstract}
  In this paper we analyze the changes in accessibility to food banks
  and related services before and during the COVID-19 pandemic in the
  City of Hamilton, Ontario. Food banks and services are the last line
  of support for households facing food insecurity; as such, their
  relevance cannot be ignored in the midst of the economic upheaval
  caused by the pandemic. Our analysis is based on the application of
  balanced floating catchment areas, and concentrates on households with
  lower incomes (\textless CAD40,000, approximately the Low Income
  Cutoff Value for a city of Hamilton's size). We find that
  accessibility was low to begin with in suburban and exurban parts of
  the city; furthermore, about 14\% of locations originally available in
  Hamilton closed during the pandemic, further reducing accessibility.
  The impact of closures on the level of service of the remaining
  facilities, and on accessibility, was disproportionate, with
  system-wide losses exceeding 40\%. Those losses were geographically
  and demographically uneven. While every part of the city faced a
  reduction in accessibility, inner suburbs fared worse in terms of loss
  of accessibility. As well, children (age \(\le 18\)) appear to have
  been impacted the most.
  \end{abstract}

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and negative physical and mental health impact in children and adults
(Elgar et al., 2021; Jones, 2017; Ramsey et al., 2011; Seligman et al.,
2010; Stuff et al., 2004). Over at least the past four decades food
banks and related services have become an essential line of defense
against food insecurity in Canadian communities (Black and Seto, 2020;
Holmes et al., 2018; Riches, 2002; Tarasuk et al., 2020). In this
respect, Canada is not unlike numerous other wealthy countries where a
systematic dismantling of the welfare state took place in the
intervening period (Tarasuk et al., 2014).

The emergence of COVID-19, the worst public health crisis since the 1918
flu pandemic, has exposed important social and economic fault lines, and
pre-existing patterns of inequality appear to have been exacerbated.
Along several other dimensions of stress (e.g., accessibility to health
care facilities, Ghorbanzadeh et al., 2021; Pereira et al., 2021a), this
seems to be the case for food insecurity as well (Laborde et al., 2020).
In the US, for example, it has been estimated that there was an increase
of more than 30\% in household food insecurity, and more than one third
of households were discovered to be newly food insecure - meaning they
did not experience food insecurity before the pandemic (Niles et al.,
2020). In Canada, Men and Tarasuk (2021) report that about 25\% of
individuals who experienced job insecurity (a relatively common
occurrence during the pandemic), also experienced food insecurity.
Similarly, according to Statistics Canada (2020a), in the early stages
of the pandemic almost 15\% of individuals reported living in a
household that faced food insecurity; the risk of food insecurity was
substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children; likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While food banks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As a mid-size city grappling with
deindustrialization, Hamilton exhibits high rates of poverty and food
bank use. As recently as 2019, the Hamilton Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that food banks in the city recorded the highest number of
visitors in the past 29 years; a rate of increase greater than
population growth. Most troubling, approximately 40\% of all visitors
were children.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary. Although more recent
statistics on food bank use in Hamilton in 2020 and 2021 are not yet
available, data from the Daily Bread Food Bank in neighbouring Toronto
for August 2020 shows visits climbing 51\% year-over-year in that
city\footnote{https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf},
which suggests that many households in Hamilton are likely to turn to
food bank services to meet their needs.

@@

With a population of around 540,000, the City of Hamilton is the fourth
largest city in Ontario. It has historically been home to major
manufacturing industries but de-industrialization that has occurred over
the past several decades has led Hamilton to become one of the most
highly divided cities in Ontario, with a significant proportion of its
residents living at or below Canada's poverty level (DeLuca et al.,
2012; Jakar and Dunn, 2019; Latham and Moffat, 2007). In 2016, the
Hamilton Community Foundation reported that based on the Low-Income
Cut-Off, Hamilton recorded a poverty rate of 16.7\%, which was well
above the average rate of Ontario (13.7\%) and the average national rate
(12.8\%).\footnote{https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/}.
According to data from Hamilton Food Share\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf},
approximately 23,000 individuals accessed food banks in the city in
March of 2019. Within this total is 9,125 visits by children (minors up
to 18 years old), up from 8,278 the year before. Feed Ontario, the
province's largest collective of hunger-relief organizations, found that
on a per-capita basis, the level of need in the inner core of central
Hamilton was second highest in Ontario\footnote{https://feedontario.ca/research/report-hunger-map/}.

Geographically, the ``old'' city of Hamilton was amalgamated with
several of its surrounding municipalities in 2001, with the city now
featuring a mix of urban, suburban, exurban, and rural areas. Lower-cost
housing proximate to the city's industrial north end has traditionally
attracted immigrants and less-affluent residents compared to the city's
wealthier suburbs. However, the decentralization of population from the
inner core has led to challenges in transit connectivity to amenities
and services and the proportion of auto users compared to transit users
remains very high (Behan et al., 2008; Topalovic et al., 2012). In
addition, the city is separated geographically by the Niagara
Escarpment. With sections of rocky cliff that approach 100m in height,
the escarpment presents a significant challenge for promoting active
travel and transport connections between ``mountain'' and ``lower city''
neighbourhoods. Taken together, the high level of food need, population
locations, and transportation network characteristics combine to inform
spatial accessibility to food banks and food services in the city.

@@

The locations of food banks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. Table 1 defines each
service type and the number of locations pre- and during the COVID-19
pandemic. While some food bank services have a specific a target
population, such as prioritizing families with young children aged
between 0 and 3 or accepting only those providing proof of low-income
status through housing and utility statements, all the food bank
services indicated below are designed to accommodate those in need of
food at zero to low cost. With our focus on food banks and food services
that offer free or low-cost meals at particular locations, we first
removed services such as Meals on Wheels and other food access services
such as food box, community kitchens, student nutrition programs, and
shopping and transportation. With some providers offering different food
services at the same location (e.g.~food bank with free and community
meal services), and some of these services closing after the onset of
the COVID-19 pandemic, we opted to geocode based on the service type. On
the other hand, two free meal services held on different days at the
same location were collapsed into a single service point for the
accessibility analysis. Additional details on the operations of
individual facilities is not publicly available, and with the changes in
operations it proved unfeasible to collect it. For this reason, the
analysis to follow is of accessibility to the location of food banks and
services, but not to specific services (e.g., breakfasts vs.~food
boxes).

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: while increases in
accessibility that result from decreases in the travel time window might
seem counter-intuitive, this is a result of lower \emph{congestion},
since fewer households are serviced and thus competition for the same
resources is more limited. System-wide accessibility declines with
higher values of \(\delta\): as more households are serviced, congestion
grows and the level of service declines, although this happens at a
declining rate. We are not aware of any research that explains how long
people are expected to travel for food banks, but we note that in
developing countries, accessible sources of drinking water are those
that can be reached in less than 30 minutes (round trip, see UNICEF-WHO,
2019). There is no reason why people in affluent countries should be
expected to spend more time travelling for a basic necessity such as
food. Accordingly, we adopt a 15-minute threshold for the analysis
(representing a one-way trip). This threshold is also approximately
where the rate of change in accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially
during the pandemic, with accessibility dropping on average by almost
41\%, but with large variations: some zones experienced changes in
accessibility of only about 10\%, whereas the most affected zone saw a
loss of accessibility of almost 71\%. Figure
\ref{fig:accessibility-changes-with-local-i} shows the changes in
accessibility. Every zone is worse off after the closure of facilities
during the pandemic, but some parts of the city seem to have been
particularly affected. To better highlight these changes, we used a
local indicator of spatial autocorrelation (Anselin, 1995) to explore
the pattern of change in accessibility. Twenty-four TAZ are flagged as
having significantly large losses of accessibility (at \(p\le 0.10\),
without correcting for multiple comparisons). Those zones are
highlighted in the figure, where it can be seen that they form more or
less compact neighborhoods. Remarkably, the largest significant drops in
accessibility are not downtown, but located in two cases in the
industrial north of the city, in one case in an inner suburb above the
escarpment, and lastly in a more suburban/exurban region in the
south-west.

@@

Food insecurity is a significant issue for many Canadian households and
is associated with reductions in nutrition and poor mental and physical
health outcomes. While food bank use can provide some relief, the levels
of need in the City of Hamilton were on the rise before the onset of the
COVID-19 pandemic, with a troubling increase in the number of visits
made by children. While temporary financial relief measures may have
helped reduce food bank use in the summer of 2020, the pandemic has in
all probability increased food insecurity for many households. To
compound matters, the pandemic has also resulted in major disruptions,
including to employment, mobility alternatives, and to food bank and
food service delivery. In response, this research has sought to better
understand accessibility to food banks and food service locations, as
well as how the closure of some locations over the pre- and during-COVID
time periods affected the potential for low-income households to reach
these amenities. Using the balanced floating catchment area method to
account for population demand and congestion effects at service points,
we estimated multi-modal accessibility to food bank and food service
locations for low-income households. The weighting of travel time
estimates by the modal split in different zonal geographies tailors the
results to patterns of travel behaviour captured in the regional travel
survey. Moreover, our parcel-level analysis presents a disaggregate
approach to estimating accessibility based on the locations of
residential parcels and dwellings.

Our results show that while accessibility levels were lower in the
city's more car-oriented suburban and rural areas to begin with, the
closure of 14\% of the city's food bank and low-cost or free food
service locations during the pandemic resulted in an overall decrease in
accessibility across the city. However, these effects were not uniform
over space or for different population groups. Since the balanced
floating catchment area method takes into account changes in demand and
congestion for service providers, the closure of food bank and food
service locations can reverberate throughout the catchment areas of the
whole city. For some suburban zones, the closure of a relatively high
level-of-service location results in the remaining services being spread
over a larger population. In others, high auto dependence for trips
leads to decreases in accessibility that accumulate due to the loss of
several food bank and service locations initially reachable within 15
minutes by car. Reductions in accessibility in the city's more urban
north end, where low-income households conduct higher proportions of
trips by transit and walking, emphasize the importance of geographic
proximity in the potential to reach food bank and service locations for
these residents. Beyond geography, the results also highlight the
differential impact of food banks closing during COVID-19 on population
groups, with seniors and children most affected.

In the absence of information regarding how food insecure households
travel to food banks and related services, we examined accessibility to
food banks using a 15 minutes (one way) travel time threshold. The fact
that we must rely on a standard created for accessible drinking water in
the developing world only serves to highlight the tragedy of food
insecurity in an affluent country like Canada. More broadly, it points
to the absurd need to understand how a bad situation was made worse by
the pandemic: in effect, the analysis reveals that disparities in
accessibility to food banks and services predated the pandemic, that the
pandemic contributed to the deterioration of these services, and that
populations already in distress, particularly children, ended up in an
even more adverse state. How much worse, it is impossible to say, mainly
because there is also a dearth of information, let alone standards,
regarding what is an \emph{acceptable} or \emph{sufficient} level of
service when it comes to food banks. Whether this is one facility per
1,000 population or three facilities per 1,000 population, or any other
number, we know not.

In summary, while previous work has noted the important role of
geography alongside other socio-economic and demographic indicators in
household access to healthy food, this is, to the best of our knowledge,
the first study that has focused on the geographic component of food
bank accessibility. Beyond accounting for the inflation of demand that
occurs in traditional floating catchment area methods, our application
of the balanced floating catchment area approach offers a novel analysis
of food bank and food service accessibility that is sensitive to the
locations of low-income households, information on their age
distributions and typical trip-making behaviour, food bank and food
service locations and their operations over time, and the
characteristics of the city's multi-modal transportation network.
Despite the limitations noted above, the findings from this analysis can
provide quantitative evidence to help urban planners and community
administrators strategically allocate food bank services considering
accessibility for different population groups, especially when
considering the unequal distribution of impacts from the COVID-19
pandemic.

In terms of future research, it is important to note that the degree to
which our low-income cut-off of CAD\$40,000 reflects food insecurity in
the different zones of the study area is not known. We also consider all
food banks and food service locations equally. Information on
differential capacities at food provider locations is currently not
collected, and given the closure of facilities was not possible to
obtain. Data on services, such as the number of meals served, could be
used to refine the analysis in the future. Moreover, while the travel
survey allows us to model multi-modal accessibilities that align with
travel behaviour observed in the travel survey and capture differences
in accessibility by age categories, the use of travel survey data for
modelling food insecurity also has its limitations. Research into the
population weighting methods used in the TTS note that the survey may
under-count the lowest- and highest-income households in the survey
study region, although the magnitude of this under-counting is unknown,
and approximately 20\% of respondents to the 2016 survey did not report
their income (Rose, 2018). In that regard, the modal splits of
low-income households observed through the TTS data may not accurately
reflect the travel behaviours of food insecure populations, and our
estimates might be somewhat conservative if those households who do not
report income rely more on walking and/or transit for their mobility
needs.

@@

\leavevmode\hypertarget{ref-behan2008smart}{}%
Behan, K., Maoh, H., Kanaroglou, P., 2008. Smart growth strategies,
transportation and urban sprawl: Simulated futures for hamilton,
ontario. The Canadian Geographer/Le G{é}ographe canadien 52, 291--308.
doi:\url{https://doi.org/10.1111/j.1541-0064.2008.00214.x}

@@

\leavevmode\hypertarget{ref-delamater2013spatial}{}%
Delamater, P.L., 2013. Spatial accessibility in suboptimally configured
health care systems: A modified two-step floating catchment area
(M2SFCA) metric. Health \& Place 24, 30--43.
doi:\url{https://doi.org/10.1016/j.healthplace.2013.07.012}

\leavevmode\hypertarget{ref-deluca2012code}{}%
DeLuca, P.F., Buist, S., Johnston, N., 2012. The code red project:
Engaging communities in health system change in hamilton, canada. Social
Indicators Research 108, 317--327.
doi:\url{https://doi.org/10.1007/s11205-012-0068-y}

@@

\leavevmode\hypertarget{ref-ghorbanzadeh2021spatial}{}%
Ghorbanzadeh, M., Kim, K., Erman Ozguven, E., Horner, M.W., 2021.
Spatial accessibility assessment of COVID-19 patients to healthcare
facilities: A case study of florida. Travel Behaviour and Society 24,
95--101. doi:\url{https://doi.org/10.1016/j.tbs.2021.03.004}

@@

\leavevmode\hypertarget{ref-jakar2019turning}{}%
Jakar, G.S., Dunn, J.R., 2019. (Turning rust into gold?) Hamilton,
ontario and a canadian perspective of shrinking and declining cities.
Cities 94, 1--10. doi:\url{https://doi.org/10.1016/j.cities.2019.05.016}

@@

\leavevmode\hypertarget{ref-latham2007determinants}{}%
Latham, J., Moffat, T., 2007. Determinants of variation in food cost and
availability in two socioeconomically contrasting neighbourhoods of
hamilton, ontario, canada. Health \& Place 13, 273--287.
doi:\url{https://doi.org/10.1016/j.healthplace.2006.01.006}

@@

\leavevmode\hypertarget{ref-topalovic2012light}{}%
Topalovic, P., Carter, J., Topalovic, M., Krantzberg, G., 2012. Light
rail transit in hamilton: Health, environmental and economic impact
analysis. Social Indicators Research 108, 329--350.
doi:\url{https://doi.org/10.1007/s11205-012-0069-x}
